RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-12-01 manuscripts/manuscript.Rmd
A few edits in response to @hlapp's suggestions

A detailed account of how the NeXML standard addresses these and other
relevant challenges can be found in @Vos_2012. In brief, NeXML was
designed with the following important properties. First, NeXML is defined
by a precise grammar that can be programmatically **validated**. This
ensures that any NeXML file can be expected to be parsed and read
without errors by any NeXML-compliant software, provided the file
passes validation. Second, NeXML is **extensible**: a user can define
representations of new, previously unanticipated information (as we
will illustrate) without violating its defining grammar. Third and
most importantly, NeXML is **semantically rich**.
The use of formal semantics in NeXML makes its data representation
not only machine-readable in a predictable manner, but also enables machine reasoning about
the data and metadata (see @Parr2011 for an overview on semantic reasoning in 
evolutionary informatics). Instead of asking the computer to
return "the data on line #93" <!-- this is a pretty obscure example, -->
<!-- and I'm not even clear what you mean here, so having a hard time -->
<!-- coming up with an improvement -->, we can expect the computer to reason about
taxonomic concepts (as we illustrate in the supplement).

@@

```{r}
nex <- nexml_read("components/trees.xml")
typeof(nex)
```

The method `nexml_read()` reads NeXML files, either from a local
file, or from a remote location via its URL, and returns an object of
type `nexml`. This object type is a representation of the NeXML
document contents, and therefore holds phylogenetic trees as well as 
character or trait matrices, and all metadata. This is similar in
design to the classes in the `phylobase` package, which contain 
trees and traits but hold only limited metadata [@phylobase], but
different from R object types such as `ape::phylo` (the `phylo` class
in the `ape` package [@Paradis2004]), which represents trees alone.

@@

The `get_trees_list()` method is designed for use in scripts, for
which a consistent and predictable return type (namely
`ape::multiPhylo`) is desirable.
<!-- the open (and previously not covered) issue here is what does -->
<!-- get_trees_list() return in the case of multiple trees blocks? HL -->
Following a common design practice
for R packages, `RNeXML` also contains a method `get_trees()`, which
is designed for interactive use and returns the R object most
intuitive given the arrangement of phylogeny data in the source NeXML
document. NeXML inherits from NEXUS the ability to contain multiple
phylogenies, and to group sets of phylogenies into "blocks". The
`get_trees()` method will return an `ape::phylo` object if the NeXML
document contains a single phylogeny, an `ape::multiPhylo` object if
it contains multiple phylogenies arranged in a single `trees` block,
and a list of `ape::multiPhylo` objects if it contains multiple
`trees` blocks. This design spares an interactive user from having to
type unnecessary data conversion commands, at the expense of needing
to know which return type to expect (or verifying the type via
introspection commands such as `typeof()`).

@@

```{r}
phy <- get_trees(nex)
typeof(phy)
plot(phy[[1]])
```

@@

To allow strict validation of data types, the NeXML schema requires
continuous and discrete character data to be separated into different
"characters" blocks. Since the `data.frame` data structure in R has no
such constraints, the `get_characters()` method will combine such
blocks as separate columns into a single `data.frame` object, provided
they correspond to the same taxa.
<!-- does the example data file actually demonstrate this? if not, it -->
<!-- should. HL -->
Otherwise, a list of `data.frame`s
is returned, with list elements corresponding to characters
blocks. Similar to the distinction between `get_trees()` and
`get_trees_list()`, there is also a method `get_characters_list()`,
which always returns a list of `data.frame`s.

@@

<!-- Might be worth adding some brief explanation here as to what -->
<!-- happens internally in such a case. For example, automatic -->
<!-- generation of multiple `otus` blocks? HL --> 
Other than storage size, there is no limit to the number of
phylogenies and character matrices that can be included in a
single NeXML file. This allows, for example, to capture samples from a
posterior probability distribution of inferred or simulated phylogenies or
character states in one NeXML file.

@@

  creator = "Sibley, C. G. and Ahlquist, J. E.",
  nexml = birds)
```
<!-- the above will get fixed by CB, this is just a reminder and -->
<!-- placeholder to adjust when that's done. -->

@@

Alternatively, a citation to a published paper with a Digital
Object Identifier (DOI) can be provided in the form of the DOI,
which the package `knitcitations` [@Boettiger_2014] can turn into a
formatted citation using. As an example, to add the citation information of the paper that generated the
`geospiza` phylogeny included in the `geiger` package:

```{r}
library("knitcitations")
geiger_nex <- add_basic_meta(citation = bib_metadata("10.2307/2408428"), nexml=geiger_nex)
```

>>>>

A detailed account of how the NeXML standard addresses these and other
relevant challenges can be found in @Vos_2012. In brief, NeXML was
designed with the following important properties. First, NeXML is defined
by a precise grammar that can be programmatically **validated**. This
ensures that any NeXML file can be expected to be parsed and read
without errors by any NeXML-compliant software, provided the file
passes validation. Second, NeXML is **extensible**: a user can define
representations of new, previously unanticipated information (as we
will illustrate) without violating its defining grammar. Third and
most importantly, NeXML is **semantically rich**.
The use of formal semantics in NeXML makes its data representation
not only machine-readable in a predictable manner, but also enables machine reasoning about
the data and metadata (see @Parr2011 for an overview on semantic reasoning in 
evolutionary informatics). For instance, through the use of semantics we can
expect the computer to reason about taxonomic concepts
(as we illustrate in the supplement). 
<!-- Sorry, just trying to contrast what it means to make a 'semantic' query 
vs a non-semantic one.  Most R users won't be familiar with SQL queries either anyway.  Let me know if this is better.-->

@@

```{r}
nex <- nexml_read("components/trees.xml")
```

The method `nexml_read()` reads NeXML files, either from a local
file, or from a remote location via its URL, and returns an object of
type `nexml`. This object type is a representation of the NeXML
document contents, and therefore holds phylogenetic trees as well as 
character or trait matrices, and all metadata. This is similar in
design to the classes in the `phylobase` package, which contain 
trees and traits but hold only limited metadata [@phylobase], but
different from R object types such as `ape::phylo` (the `phylo` class
in the `ape` package [@Paradis_2004]), which represents trees alone.

@@

The `get_trees_list()` method is designed for use in scripts, for
which a consistent and predictable return type (namely a `list` of
`ape::multiPhylo` objects, reflecting the nested structure of NeXML) is desirable.
`RNeXML` also contains a method `get_trees()`, which
is designed for interactive use and returns the R object most
intuitive given the arrangement of phylogeny data in the source NeXML
document. NeXML inherits from NEXUS the ability to contain multiple
phylogenies, and to group sets of phylogenies into "blocks". The
`get_trees()` method will return an `ape::phylo` object if the NeXML
document contains a single phylogeny, an `ape::multiPhylo` object if
it contains multiple phylogenies arranged in a single `trees` block,
and a list of `ape::multiPhylo` objects if it contains multiple
`trees` blocks. This design spares an interactive user from having to
type unnecessary data conversion commands, at the expense of needing
to know which return type to expect (or verifying the type via
introspection commands such as `class()`).

@@

```{r}
phy <- get_trees(nex)
plot(phy[[1]])
```

@@

To allow strict validation of data types, the NeXML schema requires
continuous and discrete character data to be separated into different
"characters" blocks. Since the `data.frame` data structure in R has no
such constraints, the `get_characters()` method will combine such
blocks as separate columns into a single `data.frame` object, provided
they correspond to the same taxa. Otherwise, a list of `data.frame`s
is returned, with list elements corresponding to characters
blocks. Similar to the distinction between `get_trees()` and
`get_trees_list()`, there is also a method `get_characters_list()`,
which always returns a list of `data.frame`s.

@@

`RNeXML` automatically adds new, separate OTU blocks into the NeXML file
for each of these new data sets.
Other than storage size, there is no limit to the number of
phylogenies and character matrices that can be included in a
single NeXML file. This allows, for example, to capture samples from a
posterior probability distribution of inferred or simulated phylogenies or
character states in one NeXML file.

@@

  creator = "Sibley, C. G. and Ahlquist, J. E.",
  nexml = birds)
```

<!-- the above will get fixed by CB, this is just a reminder and -->
<!-- placeholder to adjust when that's done. -->
<!-- Um, sorry, remind me what I am supposed to fix here? formatting? -->

@@

Alternatively, a citation to a published paper with a Digital
Object Identifier (DOI) can be provided in the form of the DOI,
which the package `knitcitations` [@Boettiger_2014] can turn into a
formatted citation using. As an example, to add the above citation
using the DOI:

```{r }
library("knitcitations")
birds <- add_basic_meta(citation = bib_metadata("10.1093/bioinformatics/btg412"), nexml=birds)
```
